A Type-Theoretical Wide-Coverage Computational Grammar for Swedish

### 2012-03-21 tsd/TSD2012/paper.tex
tsd updates

\begin{abstract}
The work describes a wide-coverage computational grammar for Swedish. 
It is developed using GF(Grammatical Framework), a dependently-typed 
functional language, specialized for grammar programming. We trained
and evaluated the grammar by using Talbanken, one of the largest treebanks for 
Swedish. As a result 65 $\%$ of the Talbanken trees were translated
into the GF format in the training stage and 75 \% of the noun phrases
was parsed during the evaluation. Moreover, we obtained a language
model for Swedish which we use for disambiguation when parsing free text
with the grammar. 
\keywords{grammar-formalism, computational grammar, treebank}
\end{abstract}

@@

The current work aims at overcoming the second limitation of GF
grammars --- adding support for the most common syntactic
constructions that could appear in Swedish texts. For some of these we
use dependent types to encode the syntactic phenomena in an elegant
way. The lexicon limitation is not a problem, as we use an existing 
large Swedish lexicon \cite{ahlbergEnache}, which is being continuously
developed and contains more than 120 000 entries.
%... \textbf{Malin, can you say a bit more ?But not more than a couple
%  of lines :)}

@@

% Chapter 2.4(with emphasis on the syntactic structures)
Swedish is a North-Germanic language, sharing most of its grammatical structure 
with Norwegian and Danish. It is also one of the official languages in Finland
and is spoken by approximately 9 million people.
Although Swedish syntax often is similar to English, the  morphology is richer and the
word order more intricate.
It is a verb-second language: %\cite[p.116]{gunlog}: 
the second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but fronting other constituents
-- topicalisation -- is very common,  especially for temporal and
locative adverbial phrases \cite[\textsection 1027]{H&H}.
Fronting the finite verb marks questions.
%\enumsentence{Passive \hspace{45mm} Active\\
%\shortexm{12}
%{Tjuven & togs&  av & polisen}
%{\emph{the}\emph{ thief}&  \emph{took}+\textbf{s}&  \emph{by}&\emph{the police}}
%{``The thief was arrested by the police"}
%{&&Polisen & tog & tjuven}
%{&&\emph{the police} & \emph{took} & \emph{the thief}}
%{\hspace{-9mm}``The police arrested the thief"}}
%\label{gfPass:s-pass}
Special reflexive pronouns and reflexive possessive pronouns for the 3rd
exist \cite[\textsection 310 \& 319]{H&H}, distinct from the normal 3rd person forms.
%\enumsentence{
%\begin{tabular}{llll}
%a. & Han slog \textbf{sig}.&\hspace{6mm}b. & Han såg \textbf{sitt} barn.\\
%&\emph{He hit \sc self.} && \emph{He saw {\sc self's} child.}\\
%c. &Han slog \textbf{honom}. &\hspace{6mm}d.& Han såg \textbf{hans} barn.\\
%&\emph{He hit him (another person).}&&\emph{He saw his (another person's) child.}
%\end{tabular}}

@@

%Chapter 2.2
\label{sec:talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
We have used Talbanken \cite{talbanken},
a freely available, manually annotated, large-scale treebank.
It is analyzed with the MAMBA annotation scheme (Teleman, 1974) and 
the more than 6000 sentences we have been using
%consists of four
%parts. Two of them are transcriptions of spoken language, one a collection of
%text written by high school students, and one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.

@@

%Discussion about the two dimensions of a grammar - lexicon and syntax 
There are two aspects of on a computational grammar -- the work on the lexicon
and the work regarding syntax.  For this article, we will focus on the
syntactical dimension of our work, the description and formalization of grammar
rules and the usage of Talbanken.

@@

\noindent\textbf{The s-passive}
Swedish has two ways of forming passive verb phrases: the 
periphrastic passive, formed by using the modal auxiliary verb `bli'
(\emph{`become'})
and the s-passive which is formed by adding an \emph{s} to the verb. 
Passive voice is often used, especially the
s-passive, % (sentence \ref{gfPass:s-pass}).
which is used in more than 80 \% of the times according to some studies 
%Some studies suggest that the s-passive is 
\cite{laanemets}.
It is however not as common in the other Scandinavian languages, 
where not all words have passive forms for all tenses. The resource grammar for
Scandinavian therefore only implemented the periphrastic passive.
%\verb-PassV2-, by using auxiliary verb.
%The function allows two-place verbs to be used in passive by using \emph{bli} (\emph{become}), and thereby
%turned into complete verb phrases; they no longer need an object.
During this project, the s-passive was added although the
periphrastic passive is still allowed.
%The grammar further allows not only transitive verbs, but 
%All verb phrases that misses an object are allowed to form passives.
A ditransitive verb -- \emph{ge} (`give') (\ref{sent:give2pass}) -- gives rise to two
passives, (\ref{ex:passV33}) and (\ref{ex:passV32}), both covered by our new implementation.
\eenumsentence{
\item{\textbf{Active use of two-place verb}\\\shortex{4}
{Vi & erbjöd& henne& jobbet }
{\emph{we} & \emph{offered}& \emph{her}&\emph{the job}}
{``We offered her the job"}}
\label{sent:give2pass}
\item{\textbf{Passive use, first place \hspace{3mm}Passive use, second place}\\\shortex{7}
{Hon & erbjöds & jobbet && \hspace{3mm}Jobbet & erbjöds & henne }
{\emph{she} &\emph{ offered+\textbf{s}}&\emph{ the job}
&&  \hspace{3mm}\emph{the job} & \emph{offered+\textbf{s}}&\emph{ her}}
{``She was offered the job"}
~``The job was offered to her"}}
\label{ex:passV33}
%\item{\textbf{Second place in two-place verbs}\\\shortex{3}
%{Jobbet & erbjöds & henne }
%{``The job was offered to her"}}
\label{ex:passV32}}

\subsubsection{Impersonal constructions}
\label{sec:Formal}
Formal subjects \cite[\textsection 19]{SAG} are often used in Swedish.
\enumsentence{\shortex{6}
{Det & sitter & en & fågel & på & taket}
{\emph{it }& \emph{sits }& \emph{a} & \emph{bird} &\emph{ on} & \emph{the roof}}
{``There is a bird sitting on the roof''}}
%\emph{`Det'} has the position of the subject, and the real subject, 
%\emph{`en fågel'} the one of an object.
Restrictions on both the verb and the noun phrase are covered by the grammar:
the determiner of the
noun phrase must be of such type that it requires both the noun and its modifiers
to appear in indefinite form. The verb phrase must consist of an intransitive verb
or be in passive form.
%A very common example of impersonal constructions is sentences with the verbs \emph{finnas},
%(\emph{exist})
%\emph{saknas} (\emph{miss}) and \emph{fattas}(\emph{lack}).
%\enumsentence{\begin{tabular}[t]{@{}*{18}{l@{\ }}}
%a. &Det & finns & kaffe &&&
%b. &Det & saknas&  kaffe &&&
%c. &Det & fattas & kaffe \\
%&\emph{it }& \emph{exist }& \emph{coffee} && &
% &\emph{it }& \emph{misses} & \emph{coffee} && &
% &\emph{it} & \emph{lacks} & \emph{coffee}\\
%\end{tabular}\\
%{\indent~~''There is coffee"~~~~ ''There is no coffee"~~~~ ''There is no coffee"}}

@@

In the standard GF analysis, which is preformed bottom-up starting from the
POS-tags, information about semantic roles are given by the functions, not by the
categories. For example, the first argument of the function 
\verb-PredVP- always acts as the subject, but the noun phrase itself does not carry
information about its semantic role.
As noun phrases containing reflexive pronouns may be used as ordinary
noun phrases, apart from the restrictions stated above,
we do not want them to differentiate
from other \verb-NP- on the type level since we would then need to duplicate
our code.
%due to the rich type system in GF. 
Still, the type system should prevent noun phrases containing reflexive
pronouns from being used as subjects.
This does not only concern noun phrases, but also adverbial and adjectival phrases 
(\ref{ex:ReflAP}).
%\eenumsentence{
%\item{
%taket på sitt hus\\
%\emph{the roof of {\sc self's} house}}
%\item{
%längre än sin kompis\\
%{\emph{taller than {\sc self's} friend}}}}
%\label{ex:adjTyp}

@@

\subsubsection{Overgenerating}
Another aspect of the grammar implementation was the avoidance of overgeneration. As
the Swedish resource grammar had not been used for larger projects, many examples of
overgeneration were present that did not cause problems when working with small
lexicons and controlled languages, but that caused explosions of ambiguities when 
tested on Talbanken. For the function \verb-GenNP-, used for genitive, we managed
to decrease the number of parse trees for problematic, although short, sentences to 10 \%.
% Could also be here:
%Experiments were also conducted on how we can use Talbanken for assigning correct
%types to words which we couldnot import directly from SALDO, due to annotational
%differeneces.

@@

Talbanken contains much valuable information about phrase structures and word 
usage.
One part of this project has focused on translating trees from Talbanken to GF
by constructing a mapping, which automatically transforms trees in the
Talbanken format to GF abstract syntax trees. We hence get a comparison 
%\vspace{-9mm}
\begin{figure}[!h]
\centering
\subfloat[Talbanken tree]
{\begin{tikzpicture}
\Tree [.S [.SS [.POOP Det ]]
          [.FV [.AVPS \"ar ]]
          [.NA [.ABNA inte ] ]
          [.SP [.AJ l\"att ]]
          [.IP [.IP . ] ]
      ]
\end{tikzpicture}
}
\subfloat[Translation to GF]
{\includegraphics[width=60mm]{latt2.pdf}}
\caption{The sentence \emph{Det är inte lätt} (``It is not easy"). If the word
         \emph{lätt} (`easy') is unknown to the grammar, a meta variable is put
         in its place.}
\label{fig:gftree1}
\vspace{-5mm}
\end{figure}
between the two annotations and at the same time we get methods for extracting
and translating to GF notation. 
If a node fails to be translated, a meta variable  --annotated \verb-?- -- is put in its place.
This is also the case when two sister nodes can be translated, but no rule is
applicable for joining them into one tree.

Talbanken05 uses three kinds of tags: categories, edge labels and POS-tags. 
While the POS-tags are reserved for words, the categories give grammatical information,
such as \verb|S|, \verb|NP| and \verb|VP|.
The edge labels give the part of sentence: \verb|SS| for subject, 
\verb|OO| for object etc. 
%
%\begin{tabular}{lll}
%\textbf{Tag} & \textbf{Example} & \textbf{Explanation} \\
%\hline
%\textbf{SV} & \emph{ska} (`shall') & The verb \emph{shall}\\
%\textbf{NNDDHSGG} & \emph{familjemedlemmarnas} & Noun, definite, compound (person),\\
%&(`the family members' ') &genitive \\
%\textbf{XX} & & Unclassifiable part-of-speech\\
%\end{tabular}\\
The mapping could in some cases be performed tag-by-tag, but annotational
differences complicated the translation.
One example is valency information, which is given implicitly by the complements
of a word in Talbanken. If a verb is followed by an object, \verb-OO-, which contains an
\verb-S-, we can conclude that this is a sentence-complement verb.
In GF, the valency information is given for each entry in the lexicon.
A sentence-complement verb has the type \verb-VS- and must the 
function \verb-ComplVS- must be used for combining the verb and with a 
sentence.
%Another annotational difference that turned out to be problematic for the
%translation can be illustrated by analysis of the generic pronoun \emph{`man'}.
%In Talbanken, \emph{`man'} is simply a personal pronoun, 
%in GF it is represented by using the function \verb|GenericCl| which turns a
%\verb-VP- into a clause and has no subject visible in the abstract tree. 
%When translating a sentence like this, it is thus not possible to simply glue the
%subject to the verb with \verb|PredVP|. For each similar GF construction, an extra rule
%would be needed to get full coverage.

@@

\section{Chunk Parsing}
\label{sec:chunk}
%New section - insights about latest work
The grammar, which is hand written, cannot be expected to cover all
possible expressions of the language.
Hence, we combine it with statistical means to get good coverage. 
%The parser uses named entity recognition and a compounding analysis.
The parser should give as much output as possible 
%Further we need to be able to give output
even when encountering unknown words and grammatical constructions 
or idioms and ellipses.
Our approach makes use of the rich annotation in Talbanken -- we perform chunk parsing
relying on the tags given in the treebank. 
Consequentially, whenever a sentence is not recognized by the grammar, it is 
parsed chunk by chunk and the parse trees for each chunk is returned.
Noun phrases, adverbial phrases and adjectival phrases are considered.
Due to the differences in the annotation between Talbanken and
GF mentioned is section \ref{sec:mapping} verbs are not considered
as parsable chunks. %The GF parser depend on requires verb particles and prepositions to give
%a correct output, and which are usually not in the same Talbanken chunk. 
Therefore, our focus is on whole sentences and noun phrases. At the first stage,
we try to parse the basic sentence structure, and therefore allow the parser
to parse complicated chunks as dummy words. This way, we don't loose the 
high-level sentence analysis given by GF.
When this is done, we try to separately parse the chunks,
until as much of the sentence as possible is covered.

@@

For the extraction of a GF Treebank from Talbanken, (section \ref{sec:mapping})
the evaluation measures the numbers of meta variables in each translated tree.
%If the finite verb is untranslatable, we get meta variables for the verb itself and for the
%tense.
For all of Talbanken, we get a number of 65 \%, and if we limit our input to
uncomplicated sentences with no more than 10 words where all words are known to
our lexicon, we can reach 85 \%. 
%When translating the whole treebank, only 78 sentences
%could not be mapped at all and resulted in meta variables only.

@@

\textbf{Grammar}
The grammar should cover the most prominent Swedish features.
While statistical methods compensates for constructions not covered by
the grammar, we still aim for a even wider grammar coverage. 
As examples of constructions to be added, we mention
pronominal object shift, bare indefinites and a differentiation between
object and subject control verbs.
%\vspace{-2mm}
%\begin{itemize}
%\item
%\textit{} % is common in Swedish and obligatory in Danish and Norwegian.
%\enumsentence{\begin{tabular}[t]{@{}*{11}{l@{\ }}}
%a. & Jag & ser&  honom& \textbf{inte}&~~ & b.&Jag &ser &\textbf{inte} &honom\\
%   & \emph{I} & \emph{see}&\emph{him}&\emph{not}&&& \emph{I} &\emph{see}&\emph{not}&\emph{him}\\
%\end{tabular}}
%Personal pronouns are allowed to precede negation and some other sentence adverbials % --\textit{satsadverbial} --
%in main clauses without auxiliary verbs.
%
%\enumsentence{\begin{tabular}[t]{@{}*{12}{l@{\ }}}
%a. & *Vi & har&  honom&  inte&sett&~~ & b.& * Jag &ser &huset &inte\\
%   & \emph{~we} & \emph{have}&  \emph{him}&  \emph{not} &  \emph{seen} && b.& \emph{~I}&\emph{see} &\emph{the house} &\emph{not}\\
%\end{tabular}}
%
%Although object shifts are frequently used, they are hardly found in
%Talbanken's part P, which has been the inspiration for this project.
%Therefore, this implementation has so far not been prioritized.\\

@@

All parts of the project is open-source and the grammar and other parts
are modular and may well be reused and developed.

>>>>

\begin{abstract}
The work describes a wide-coverage computational grammar for Swedish. 
It is developed using GF(Grammatical Framework), a dependently-typed 
functional language, specialized for grammar programming. We trained
and evaluated the grammar by using Talbanken, one of the largest treebanks for 
Swedish. As a result 65 $\%$ of the Talbanken trees were translated
into the GF format in the training stage and 76 \% of the noun phrases
was parsed during the evaluation. Moreover, we obtained a language
model for Swedish which we use for disambiguation when parsing free text
with the grammar. 
\keywords{grammar-formalism, computational grammar, treebank}
\end{abstract}

@@

The current work aims at overcoming the second limitation of GF
grammars --- adding support for the most common syntactic
constructions that could appear in Swedish texts. For some of these we
use dependent types to encode the syntactic phenomena in an elegant
way. The lexicon limitation is not a problem, as we use an existing 
large Swedish lexicon \cite{ahlbergEnache}, which is imported to GF
format from an electronic dictionary and which contains more than
%being continuously
%developed and contains more than 
100 000 entries.
%... \textbf{Malin, can you say a bit more ?But not more than a couple
%  of lines :)}

@@

% Chapter 2.4(with emphasis on the syntactic structures)
Swedish is a North-Germanic language, sharing most of its grammatical structure 
with Norwegian and Danish. It is also one of the official languages in Finland
and is spoken by approximately 9 million people.
Although Swedish syntax often is similar to English, the  morphology is richer and the
word order more intricate.
It is a verb-second language: %\cite[p.116]{gunlog}: 
the second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but fronting other constituents
-- topicalisation -- is very common,  especially for temporal and
locative adverbial phrases \cite[\textsection 1027]{H&H}.
Fronting the finite verb marks questions.
%\enumsentence{Passive \hspace{45mm} Active\\
%\shortexm{12}
%{Tjuven & togs&  av & polisen}
%{\emph{the}\emph{ thief}&  \emph{took}+\textbf{s}&  \emph{by}&\emph{the police}}
%{``The thief was arrested by the police"}
%{&&Polisen & tog & tjuven}
%{&&\emph{the police} & \emph{took} & \emph{the thief}}
%{\hspace{-9mm}``The police arrested the thief"}}
%\label{gfPass:s-pass}
Special reflexive pronouns and reflexive possessive pronouns for the 3rd
person exist \cite[\textsection 310 \& 319]{H&H}, distinct from the normal 3rd
person forms.
%\enumsentence{
%\begin{tabular}{llll}
%a. & Han slog \textbf{sig}.&\hspace{6mm}b. & Han såg \textbf{sitt} barn.\\
%&\emph{He hit \sc self.} && \emph{He saw {\sc self's} child.}\\
%c. &Han slog \textbf{honom}. &\hspace{6mm}d.& Han såg \textbf{hans} barn.\\
%&\emph{He hit him (another person).}&&\emph{He saw his (another person's) child.}
%\end{tabular}}

@@

%Chapter 2.2
\label{sec:talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
We have used Talbanken \cite{talbanken},
a freely available, manually annotated, large-scale treebank.
It is analyzed with the MAMBA annotation scheme and the section used for our
work contains more than 6000 sentences
%consists of four we have been using
%parts. Two of them are transcriptions of spoken language, one a collection of
%text written by high school students, and one, section P,
of professionally written Swedish gathered from newspapers, brochures and textbooks.

@@

%Discussion about the two dimensions of a grammar - lexicon and syntax 
There are two aspects of a computational grammar -- the work on the lexicon
and the work regarding syntax.  
This article focuses on the syntactical dimension and the following section
illustrates the work on describing grammatical constructions in the GF 
formalism.
%For this article, we will focus on the
%of our work, the description and formalization of grammar
%rules and the usage of Talbanken.

@@

\noindent\textbf{The s-passive}
Swedish has two ways of forming passive verb phrases: the 
periphrastic passive, formed by using the modal auxiliary verb \emph{bli}
(`become')
and the s-passive which is formed by adding an \emph{s} to the verb. 
Passive voice is often used, especially the
s-passive, % (sentence \ref{gfPass:s-pass}).
which is used in more than 80 \% of the times according to some studies 
%Some studies suggest that the s-passive is 
\cite{laanemets}.
It is however not as common in the other Scandinavian languages, 
where not all words have passive forms for all tenses. The resource grammar for
Scandinavian therefore only implemented the periphrastic passive.
%\verb-PassV2-, by using auxiliary verb.
%The function allows two-place verbs to be used in passive by using \emph{bli} (\emph{become}), and thereby
%turned into complete verb phrases; they no longer need an object.
During this project, an implementation of   the s-passive was added
%although the periphrastic passive is still allowed.
%The grammar further allows not only transitive verbs, but 
%All verb phrases that misses an object are allowed to form passives.
A ditransitive verb -- \emph{ge} (`give') (\ref{sent:give2pass}) -- gives rise to two
passives, (\ref{ex:passV32}a) and (\ref{ex:passV32}b), both covered by our grammar.
\eenumsentence{
\item{\textbf{Active use of two-place verb}\\\shortex{4}
{Vi & erbjöd& henne& jobbet }
{\emph{we} & \emph{offered}& \emph{her}&\emph{the job}}
{``We offered her the job"}
\label{sent:give2pass}}
\item{\textbf{Passive use, first place} &c. \textbf{Passive use, second place}}
\\\shortex{6}
{Hon & erbjöds & jobbet &\hspace{18pt}Jobbet & erbjöds & henne }
{\emph{she} &\emph{ offered+\textbf{s}}&\emph{ the job}
&  \hspace{18pt}\emph{the job} & \emph{offered+\textbf{s}}&\emph{ her}}
{``She was offered the job"}
~~``The job was offered to her"}
\label{ex:passV33}}
%\item{\textbf{Second place in two-place verbs}\\\shortex{3}
%{Jobbet & erbjöds & henne }
%{``The job was offered to her"}}
\label{ex:passV32}}

\subsubsection{Impersonal constructions}
\label{sec:Formal}
Formal subjects \cite[\textsection 19]{SAG} are often used in Swedish.
\enumsentence{\shortex{6}
{Det & sitter & en & fågel & på & taket}
{\emph{it }& \emph{sits }& \emph{a} & \emph{bird} &\emph{ on} & \emph{the roof}}
{``There is a bird sitting on the roof''}}
%\emph{`Det'} has the position of the subject, and the real subject, 
%\emph{`en fågel'} the one of an object.
Restrictions on both the verb and the noun phrase are covered by the grammar:
the determiner of the
noun phrase must be of such type that it requires both the noun and its modifiers
to appear in indefinite form \cite{cooper}. The verb phrase must consist of an intransitive verb
or be in passive form.
%A very common example of impersonal constructions is sentences with the verbs \emph{finnas},
%(\emph{exist})
%\emph{saknas} (\emph{miss}) and \emph{fattas}(\emph{lack}).
%\enumsentence{\begin{tabular}[t]{@{}*{18}{l@{\ }}}
%a. &Det & finns & kaffe &&&
%b. &Det & saknas&  kaffe &&&
%c. &Det & fattas & kaffe \\
%&\emph{it }& \emph{exist }& \emph{coffee} && &
% &\emph{it }& \emph{misses} & \emph{coffee} && &
% &\emph{it} & \emph{lacks} & \emph{coffee}\\
%\end{tabular}\\
%{\indent~~''There is coffee"~~~~ ''There is no coffee"~~~~ ''There is no coffee"}}

@@

In the standard GF analysis, which is preformed bottom-up starting from the
POS-tags, information about semantic roles are given by the functions, not by the
categories. For example, the first argument of the function 
\verb-PredVP- always acts as the subject, but the noun phrase itself does not carry
information about its semantic role.
As noun phrases containing reflexive pronouns may be used as ordinary
noun phrases -- apart from the restrictions stated above --
we do not want them to differentiate
from other \verb-NP- on the type level since this would require code duplication.
%due to the rich type system in GF. 
Still, the type system should prevent noun phrases containing reflexive
pronouns from being used as subjects.
This does not only concern noun phrases, but also adverbial and adjectival phrases 
(\ref{ex:ReflAP}).
%\eenumsentence{
%\item{
%taket på sitt hus\\
%\emph{the roof of {\sc self's} house}}
%\item{
%längre än sin kompis\\
%{\emph{taller than {\sc self's} friend}}}}
%\label{ex:adjTyp}

@@

\subsubsection{Overgeneration}
Another aspect of the grammar implementation was the avoidance of overgeneration. As
the Swedish resource grammar had not been used for larger projects, many examples of
overgeneration were present that did not cause problems when working with small
lexicons and controlled language, but that caused explosions of ambiguities when 
tested on Talbanken. For the function \verb-GenNP-, used for genitive, we managed
to decrease the number of parse trees for problematic, although short, sentences to 10 \%.

% Could also be here:
%Experiments were also conducted on how we can use Talbanken for assigning correct
%types to words which we couldnot import directly from SALDO, due to annotational
%differeneces.

@@

Talbanken contains much valuable information about phrase structures and word 
usage.
One part of this project has focused on translating trees from Talbanken to GF
by constructing a mapping, which automatically transforms trees in the
Talbanken format to GF abstract syntax trees. We hence get a comparison 
\begin{figure}[!h]
\centering
\subfloat[Talbanken tree]
{\includegraphics[width=60mm]{TBTree_cropped.pdf}}
\subfloat[Translation to GF]
{\includegraphics[width=60mm]{latt2.pdf}}
\caption{The sentence \emph{Det är inte lätt} (``It is not easy"). If the word
         \emph{lätt} (`easy') is unknown to the grammar, a meta variable is put
         in its place.}
\label{fig:gftree1}
\end{figure}
between the two annotations and at the same time we extract a GF treebank.
If a node fails to be translated, a meta variable  --annotated \verb-?- -- is put in its place.
This is also the case when two sister nodes can be translated, but no rule is
applicable for joining them into one tree.

Talbanken05 uses three kinds of tags: categories, edge labels and POS-tags. 
While the POS-tags are reserved for words, the categories give grammatical information:
 \verb|S|, \verb|NP| and \verb|VP|.
The edge labels give the part of sentence: \verb|SS| for subject, 
\verb|OO| for object etc. 
%
%\begin{tabular}{lll}
%\textbf{Tag} & \textbf{Example} & \textbf{Explanation} \\
%\hline
%\textbf{SV} & \emph{ska} (`shall') & The verb \emph{shall}\\
%\textbf{NNDDHSGG} & \emph{familjemedlemmarnas} & Noun, definite, compound (person),\\
%&(`the family members' ') &genitive \\
%\textbf{XX} & & Unclassifiable part-of-speech\\
%\end{tabular}\\
The mapping could in some cases be performed tag-by-tag, but annotational
differences complicated the translation.
One example is valency information, which is given implicitly by the complements
of a word in Talbanken. If a verb is followed by an object, \verb-OO-, which contains an
\verb-S-, we can conclude that this is a sentence-complement verb.
In GF, the valency information is given explicitly for each entry in the lexicon.
A sentence-complement verb has the type \verb-VS- and the 
function \verb-ComplVS-  must be used for combining the verb and  a 
sentence into a complete verb phrase.
%Another annotational difference that turned out to be problematic for the
%translation can be illustrated by analysis of the generic pronoun \emph{`man'}.
%In Talbanken, \emph{`man'} is simply a personal pronoun, 
%in GF it is represented by using the function \verb|GenericCl| which turns a
%\verb-VP- into a clause and has no subject visible in the abstract tree. 
%When translating a sentence like this, it is thus not possible to simply glue the
%subject to the verb with \verb|PredVP|. For each similar GF construction, an extra rule
%would be needed to get full coverage.

@@

\section{Chunk Parsing}
\label{sec:chunk}
%New section - insights about latest work
The grammar, which is hand written, cannot be expected to cover all
possible expressions of the language.
Hence, we combine it with statistical means to get good coverage. 
%The parser uses named entity recognition and a compounding analysis.
The parser should give as much output as possible 
%Further we need to be able to give output
even when encountering unknown words and grammatical constructions,
idioms and ellipses.
Our approach makes use of the rich annotation in Talbanken -- we perform chunk parsing
relying on the tags given in the treebank. 
Consequentially, whenever a sentence is not recognized by the grammar, it is 
parsed chunk by chunk and the parse trees for each chunk is returned.
Noun phrases, adverbial phrases and adjectival phrases are considered.
Due to the differences in the annotation between Talbanken and
GF mentioned is section \ref{sec:mapping}, verbs are not considered
as parsable chunks. %The GF parser depend on requires verb particles and prepositions to give
%a correct output, and which are usually not in the same Talbanken chunk. 
Therefore, our focus is on whole sentences and noun phrases. At the first stage,
we try to parse the basic sentence structure, and therefore allow the parser
to parse complicated chunks as dummy words. This way, we don't loose the 
high-level sentence analysis given by GF.
When this is done, we try to separately parse the chunks,
until as much of the sentence as possible is covered.

@@

For the extraction of a GF Treebank from Talbanken (section \ref{sec:mapping}),
the evaluation measures the numbers of meta variables in each translated tree.
%If the finite verb is untranslatable, we get meta variables for the verb itself and for the
%tense.
For all of Talbanken, we get a number of 65 \%, and if we limit our input to
uncomplicated sentences with no more than 10 words where all words are known to
our lexicon, we can reach 85 \%. 
%When translating the whole treebank, only 78 sentences
%could not be mapped at all and resulted in meta variables only.

@@

\textbf{Grammar}
The grammar should cover the most prominent Swedish features.
While statistical methods compensates for constructions not covered by
the grammar, we still aim for an even broader grammar coverage. 
As examples of constructions to be added, we mention
pronominal object shift, bare indefinites and a differentiation between
object and subject control verbs.
%\vspace{-2mm}
%\begin{itemize}
%\item
%\textit{} % is common in Swedish and obligatory in Danish and Norwegian.
%\enumsentence{\begin{tabular}[t]{@{}*{11}{l@{\ }}}
%a. & Jag & ser&  honom& \textbf{inte}&~~ & b.&Jag &ser &\textbf{inte} &honom\\
%   & \emph{I} & \emph{see}&\emph{him}&\emph{not}&&& \emph{I} &\emph{see}&\emph{not}&\emph{him}\\
%\end{tabular}}
%Personal pronouns are allowed to precede negation and some other sentence adverbials % --\textit{satsadverbial} --
%in main clauses without auxiliary verbs.
%
%\enumsentence{\begin{tabular}[t]{@{}*{12}{l@{\ }}}
%a. & *Vi & har&  honom&  inte&sett&~~ & b.& * Jag &ser &huset &inte\\
%   & \emph{~we} & \emph{have}&  \emph{him}&  \emph{not} &  \emph{seen} && b.& \emph{~I}&\emph{see} &\emph{the house} &\emph{not}\\
%\end{tabular}}
%
%Although object shifts are frequently used, they are hardly found in
%Talbanken's part P, which has been the inspiration for this project.
%Therefore, this implementation has so far not been prioritized.\\

@@

All parts of the project is open-source and the grammar and other parts
are modular and may well be reused and further developed.
